---
title: 'STAT656: Style Guide'
subtitle: Introduction to Coding (in R)
output:
  html_document: default
---
# Introduction
Let's define a couple of objects off of the famous Iris data set.
```{r}
X = iris$Sepal.Length
Y = iris$Sepal.Width
```

# Naming conventions
Use expressive names and a consistent scheme for naming variables. 

I happen to prefer `Camel case', whereby objectNamesLookLikeThis
```{r}
leastSquaresFit     = lm(Y~X)
betaHatLeastSquares = coef(leastSquaresFit)
```
There are other conventions, such as 

* Pascal's case: LeastSquaresFit

* Snake case: least_squares_fit

* kebab case: least-squares-fit

Classically, R has used a convention: `least.squares.fit'  Try not to use it, as it conflicts with other languages e.g. Python

# Consistent spacing
This means:

* Indentation inside functions/loops
* spaces between object names and assignment operator (`=')
* comments should be indented to the code in which they appear (somewhat counterintuitive: limit the number of comments in your code)
```{r}
X          = iris$Sepal.Length
Y          = iris$Sepal.width
n          = length(X)
runningSum = 0

for(i in 1:n){
  #each step grabs an element in X, squares it, and adds it to the running total
  squaredEntry = X[i]**2
  runningSum   = runningSum + squaredEntry
}
```

# Simple over complex
Tend to keep functions and lines of code doing one/few things.  For functions, include some meta data as to how the function works. 

Let's look at writing a code for centering and scaling an array 
```{r}
columnScaleF = function(X){
#### columnScaleF(X)
####      Arguments: 
####               * X is an n by p matrix which we are 
####                  normalizing to have column sample
####                  standard deviation equal to 1.
  Xstd    = apply(X,2,sd)
  Xscaled = t(t(X)/Xstd)
  return( Xscaled )
}
columnCenterF = function(X){
#### columnCenterF(X)
####      Arguments: 
####               * X is an n by p matrix which we are 
####                  normalizing to have column sample
####                  mean equal to zero.
  XcenteringMatrix = apply(X,2,mean)
  Xcentered        = t( t(X) - XcenteringMatrix)
  return( Xcentered )
}
```
Ideally, these two functions would be saved in separate files called 'columnCenter.r' and 'columnScale.r'  

Then, we would have a third function

```{r}
scaleNewF = function(X,center=TRUE,scale=TRUE){
#### scaleNew(X,center=TRUE,scale=TRUE)
####      Arguments: 
####               * X is an n by p matrix which we are 
####                  normalizing to have either (or both) 
####                  column sample mean equal to zero
####                  or column sample sd equal to one.
  source(columnCenter.r)
  source(columnScale.r)
  if(center) X = columnCenter(X)
  if(scale)  X = columnScale(X)
  return( X )
}
```
